---
title: "Replicating 'The Role of Time-Varying Price Elasticities in Accounting for Volatility Changes in the Crude Oil Market' (Baumeister and Peersman, 2013)"
author: "Mauricio Vargas Sepulveda"
date: 2025-12-10
format: 
  beamer:
    theme: metropolis
    navigation: horizontal
    fontsize: 12pt
    include-in-header: dimensions.tex
bibliography: references.bib
---

# Explanation

## Motivation

* @baumeister_role_2013 provides documented datasets and codes.
* The effort here consisted in translating the MATLAB code to C++ (Armadillo) and R for a fully portable replication.
* I used @vargassepulveda_armadillo4r_2025 for a straightforward open source rewrite of the MATLAB code.

## Technique and Data

* Time-varying parameter VAR (TVP-VAR-SV) to model the crude oil market because:
* Suitable for the oil sector where new production in response to price incentives will be gradual and subject to capacity constraints.
* Efforts toward other energy sources are also reflected in slow variation in the energy share over time.
* Burns = 50,000; Draws = 5,000

**Code and reorganized data**: \url{https://github.com/pachadotdev/armadillo4r-replication}

# Replication of Fig 2. Time-varying median impact impulse responses

```{=latex}
\begin{columns}[T]
  \begin{column}{0.2\textwidth}
    {\scriptsize
    \begin{itemize}
      \item Dark blue lines:\\Median response
      \item Shades of blue areas:\\68\% and 95\% posterior credible intervals resp.
      \item Black lines:\\Range of admissible models
    \end{itemize}
    }
  \end{column}

  \begin{column}{0.8\textwidth}
```

```{r fig2, echo=FALSE, warning=FALSE, message=FALSE, fig.width=6, fig.height=5.2, out.width='\\linewidth'}
library(ggplot2)
dinp <- readRDS("results-burns50000-draws5000.rds")
dinp$figure2
```

```{=latex}
  \end{column}
\end{columns}
```

# Rep. of Fig 3. Spectrum of price inflation and Fig 4. Short-run price elasticities

```{=latex}
\begin{columns}[T]
  \begin{column}{0.2\textwidth}
    {\scriptsize
    For Fig 3 (top panel):
    \begin{itemize}
      \item Dark blue line:\\Median response
      \item Shade of blue area:\\95\% posterior distribution.
    \end{itemize}

    \vskip0.5cm

    For Fig 4 (bottom three panels):
    \begin{itemize}
      \item Dark blue line:\\Median response
      \item Shade of blue areas: \\68\% and 95\% posterior credible intervals resp.
      \item Black lines:\\Range of admissible models
    \end{itemize}
    }
  \end{column}

  \begin{column}{0.8\textwidth}
```

```{r fig3-4, echo=FALSE, warning=FALSE, message=FALSE, fig.width=6, fig.height=5.2, out.width='\\linewidth'}
library(patchwork)
# match styles
g1 <- dinp$figure3 + theme(plot.title = element_text(face = "plain", size = 9))
g2 <- dinp$figure4
g1 / g2 + plot_layout(heights = c(1, 3))
```

```{=latex}
  \end{column}
\end{columns}
```

# Discussion

* Replication effort: Medium but for a good reason (elegant model).
* Results: Identical to the original paper.
* The article and supplentary material allow to match the results, and therefore an extension of this would be to conduct a complete replication.
* The testing model (B=5,000; D=500) took around 30 minutes to run to then correct and test again until matching the results.
* The full model as in the article (B=50,000; D=500) took around 2 hours to run to then correct and test again until matching the results.
* One major scalability drawback is the computational cost of Bayesian estimation with MCMC.

# References

## References

\footnotesize
